---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tsExtremes

<!-- badges: start -->
<!-- badges: end -->

The package tsExtremes aims to compile functions for assessing the time dependencies of extremes.

## Installation

<!-- You can install the released version of tsExtremes from [CRAN](https://CRAN.R-project.org) with: -->

You can install the package from github using

``` r
devtools::install_github('GBuritica/tsExtremes')
```

## Example

Please see below a classical data analysis using the tsExtremes package.

```{r example}
library(tsExtremes)
## basic example code
```

The package includes the daily rainfall amounts data set of weather stations from nine different locations in France.

```{r rainfall}
head(rainfall)
```

For example, the (tail) index Hill Plot of summer rainfall in Brest is ploted below.

```{r hill, echo = T,fig.height = 4, fig.width = 9, fig.align = "center"}
h <- hillestimate(rainfall$BREST[rainfall$SEASON=="SUMMER"], plot=T)
## vector of Hill estimates
```
