---
title: "Tail index inference"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This tutorial reviews statistical methods from extreme value theory for risk assessment. 
Quantifying risk is crucial in numerous engineering, econometric or climate science applications to prevent the societal, economic and environmental consequences of hazards.
For example, hydrologists carefully analyse extreme precipitation periods because these can lead to flooding or land sliding, and for this reason, their reports are essential for policymakers. 
For example, to determine the largest daily precipitation amounts that will be recorded in the next 20 years we can use the tools is extreme value theory. 


First, we start by loading the package. 
```{r setup}
library(tsExtremes)
```



## Hill estimation

To get started with the tsExtremes package, we have included a data set of Precipitations. The rainfall variable contains daily rainfall records from nine different weather stations in France. The data and season of each record is available. To compute the (tail) index Hill Plot of summer rainfall in Brest we use the function below.

```{r hill, echo = T,fig.height = 4, fig.width = 9, fig.align = "center"}
h <- hillestimator(rainfall$BREST[rainfall$SEASON=="SPRING"], plot=T)
h <- alphaestimator(rainfall$BREST[rainfall$SEASON=="SPRING"], plot=TRUE , R0 = 100,  hill=T ,  k1 = 300 )
```
