---
title: "Section 1: Tail-index inference"
author: Gloria Buriticá
geometry: margin=3cm
bibliography: "references.bib"
---

<style>
p.comment{
background-color: #E3E7F1;
padding: 10px;
border: 1px solid black;
margin-left: 25px;
border-radius: 5px;
font-style: italic;
}



</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Main goals of this tutorial
  1. Review heavy-tailed random variables.
  2. Compute estimates of the tail index.


The tail-index $\alpha > 0$ of a heavy-tailed random variable $X$ is associated with the magnitude of rare events, in this section we review a definition of heavy-tailed variables.
The tail-index is a key statistic in risk assessment applications because it measures the magnitude of extremes. Roughly speaking, when $\alpha$ is small, the extremes are highly significant, and from a practical perspective, this translates to high-impact consequences.  For example, in hydrology heavy precipitation amounts lead to devastating losses.
To sum up, if $\alpha$ is small, extremely high records are more likely to happen than if $\alpha$ is large. 


### Heavy-tailed random variables

In our setting, we consider a real-valued random variable $X$ and we say it is heavy-tailed when its tail distribution has a polynomial decay. 
To fomalize this definition we introduce the concept of regular variation.
<br>
<br>

<p class="comment">
**Definition.** (Regular variation)
We say that a real valued random variable $X$ with distribution $F$ is regularly varying with **tail-index** $\alpha > 0$ if $\lim_{t \to \infty}\bar F(tx)/\bar F(t) = x^{-\alpha},$ where $\bar F(t) := 1- F(t)$ is the tail distribution.
<p>

In the following we call heavy-tailed variables those with regularly varying distributions.

__Examples of heavy-tailed variables__ are the Pareto random variables, the Fréchet, the $t-$student, the Burr or the log Gamma distribution. 

__Examples of light-tailed distributions__ include then the exponential, the Weibull and the gaussian distribution. For these light-tailed examples, its tail distributions decay at an exponential rate and not at a polynomial rate as for regularly varying distribution. In practice this means that is very rare to record high levels sampling from them, and instead the values concentrate around the mean. 


#### Moments of regularly varying distributions
In the case of regularly varying random variables, the variance and the mean can sometimes equal infinity. This means that sampling from this variable, we obtain highly scattered observations reaching higher records in comparison with light-tailed distributions. To see this consider the following samples, and its respective boxplot.

```{r scattered observations, echo = T, message=F, fig.height = 2, fig.width = 4, fig.align = "center"}
library(VGAM)
library(ggplot2)
n     <- 500
alpha <- 2
sample_exponential <- 1 + rexp(n)
sample_pareto      <- VGAM::rpareto(n, shape=alpha)
```

```{r scattered observations plot, echo = F,fig.height = 2, fig.width = 4, fig.align = "center"}
df <- data.frame( values = c(sample_exponential,sample_pareto), type = c(rep('Exp',n), rep('Pareto',n)) )
p <- ggplot(df) + geom_boxplot(aes(y=values, color = type)) 
p + theme(
  plot.title  = element_text('Comparison light- and heavy-tailed samples'),
  axis.title.x = element_blank(),
  axis.text.x = element_blank(),
  axis.ticks  = element_blank()
  ) ## code for styling margins
```

As we already mentioned, the tail index of a heavy-tailed random variable plays an important role in determining the heaviness of tails. 

__Moments' property: __ If $X$ is non-negative and regularly varying with index $\alpha > 0$, then for $p > \alpha$, $E[X^p] = + \infty$, and for $p < \alpha$, $E[X^p] < + \infty$.

In particular the moments' property entails that for $\alpha \in (0,2)$, the variance of the variable $X$ is infinite and for $\alpha \in (0,1)$ its mean is infinite. In other words, small values of $\alpha$ indicate heavier tails than larger values of $\alpha$.
For these reasons, estimating the tail-index of a random variable $X$ is important to quantify the magnitude of rare events. 
For a full overview on regularly varying distributions we refer to  [@bingham:goldie:teugels:1987].




### Inference of the tail-index

In this subsection we now turn to the problem of esimating the tail-index from data.
We consider a sample of observations $X_1, \dots, X_n$ with the same distribution of $X$, and not necessarily independent.

<p class="comment">
**Hill estimator.**
The Hill estimator is a general procedure to estimate the tail-index of $X$, and it is given by 
$$
\frac{1}{\widehat{\alpha}^n} \;  := \;  \frac{1}{\widehat{\alpha}^n(k^\prime)}  := \frac{1}{k^\prime} \, \sum_{t=1}^{n} \log(X_t/X_{(k^\prime+1)})1(X_t > X_{(k^\prime +1)}),
$$
where $X_{(1)} \geq X_{(2)} \geq \dots \geq  X_{(n)}$ are the sorted sample records, and $k^\prime$ is a tuning parameter for the Hill estimator.
<p> 


When the temporal dependence is not very strong, think for example in a stationary AR(1) series: $(X_t = \rho X_{t-1} + Z_{t})$, for $|\rho| < 1$, and $(\mathbb{Z})$ a series of iid regularly varying innovations, the Hill estimator yields consistent estimates of the tail-index of $X$. 
We refer to Chapter 4 in [@beirlant:goegebeur:teugels:segers:dewaal:ferro:2004] for a detailed treatment of tail-index inference.


To motivate the Hill estimator, note that if $X$ follows the Pareto distribution with tail index $\alpha$, then $F(t) = 1 -x^{-\alpha}$, and then straightforward calculations imply  $E[\log(X)] = 1/\alpha$. 
In all generality, the renormalized largest records of the sample follow asymptotically a Pareto distribution by means of the Pickands-Balkdema de Haan Theorem also called 'Peaks over Thresholds' Theorem. 
This means, that if we restrict our sample to the $k^\prime$ largest records, then we see how the Hill estimator is a good candidate to estimate tail-index.
We refer in all generality to Chapter 4 in [@beirlant:goegebeur:teugels:segers:dewaal:ferro:2004].

### tsExtremes package implementation
\par 
To implement the Hill estimator with the tsExtremes package,  we start by loading the $\texttt{tsExtremes}$ package.

```{r setup}
library(tsExtremes)
```


The $\texttt{tsExtremes}$ package includes the rainfall dataset, containing daily rainfall records from nine different weather stations in France. The data and season of each record is available. To compute the tail-index and produce a Hill Plot of summer rainfall in Brest we use the function below.



```{r unbiased-hill, echo = T,fig.height = 4, fig.width = 9, fig.align = "center"}
h <- alphaestimator(rainfall$BREST[rainfall$SEASON=="SPRING"], plot=F , R0 = 100,  hill=T ,  k1 = 300 )
```


## References
