---
title: "Tail index inference"
author: Gloria Buriticá
geometry: margin=3cm
bibliography: "/Users/Buritica/Dropbox/These/git/tsExtremes/references.bib"
---

<style>
p.comment{
background-color: #E3E7F1;
padding: 10px;
border: 1px solid black;
margin-left: 25px;
border-radius: 5px;
font-style: italic;
}



</style>

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Main goals of this tutorial
  1. Review heavy-tailed random variables.
  2. Compute estimates of the tail index.


The tail index $\alpha > 0$ of a heavy-tailed random variable $X$ is associated with the magnitude of rare event. 
If $\alpha$ is small, high records are more likely to happen than if $\alpha$ is large.  


### Heavy-tailed random variables

In our setting, a real-valued random variable $X$ is heavy-tailed when its tail distribution has a polynomial decay. To state this condition formally, we will need the notion of regular variation. 
<p class="comment">
**Definition.** (Regular variation)
A real valued random variable $X$ with distribution $F$ is regularly varying with **tail-index** $\alpha > 0$ if $\lim_{t \to \infty}\bar F(tx)/\bar F(t) = x^{-\alpha},$ where $\bar F(t) := 1- F(t)$ is the tail distribution.
<p>

In the following we call heavy-tailed variables those with regularly varying distributions. __Examples__ of heavy-tailed variables are the Pareto random variables, the Fréchet, the $t-$student, the Burr or the log Gamma distribution. 
In comparison, light-tailed distributions include the exponential, the Weibull and the gaussian distribution. 
For these light-tailed examples, its tail distributions decay at an exponential rate. 
In practice this means that is very rare to record high levels sampling from them, and instead the values concentrate around the mean. 
For heavy-tailed random variables, the variance and the mean can sometimes equal infinity as we see next.
\par 
As we already mentioned, the tail index of a heavy-tailed random variable plays an important role in determining the heaviness of tails. Moreover, if $X$ is non-negative and regularly varying with index $\alpha > 0$, then for $p > \alpha$, $E[X^p] = + \infty$, and for $p < \alpha$, $E[X^p] < + \infty$. 
In particular this means that for $\alpha \in (0,2)$, the variance of the variable $X$ is infinite and for $\alpha \in (0,1)$ its mean is infinite. 
In other words, small values of $\alpha$ indicate heavier tails than larger values of $\alpha$. 
For these reasons, estimating the tail-index of a random variable $X$ is important to quantify the magnitude of rare events. 
For a full overview on regularly varying distributions we refer to  [@bingham:goldie:teugels:1987].

### Inference of the tail-index

To estimate the tail-index, we consider a sample of observations $X_1, \dots, X_n$ with the same distribution of $X$, and not necessarily independent.
For weakly dependent stationary time series, think for example we sample from an AR(1) model, or from an independent series. 
In these case, the Hill estimator is a general procedure to estimate the tail-index of $X$, and it is given by 
$$
\frac{1}{\widehat{\alpha}^n} \;  := \;  \frac{1}{\widehat{\alpha}^n(k^\prime)}  := \frac{1}{k^\prime} \, \sum_{t=1}^{n} \log(X_t/X_{(k^\prime+1)})1(X_t > X_{(k^\prime +1)}),
$$
where $X_{(1)} \geq X_{(2)} \geq \dots \geq  X_{(n)}$ are the sorted sample records, and $k^\prime$ is a tuning parameter for the Hill estimator.
Actually, if $X$ follows the Pareto distribution with tail index $\alpha$ then $F(t) = 1 -x^{-\alpha}$, and the strightforward calculations imply  $E[\log(X)] = 1/\alpha$. 
In all generality, the renormalized largest records of the sample follow asymptotically a Pareto distribution by means of the Pickands-Balkdema de Haan Theorem also called 'Peaks over Thresholds' Theorem. 
This means, that if we restrict our sample to the $k^\prime$ largest records, then we see how the Hill estimator is a good candidate to estimate tail-index.
We refer in all generality to Chapter 4 in [@beirlant:goegebeur:teugels:segers:dewaal:ferro:2004].

### tsExtremes package implementation
\par 
To implement the Hill estimator with the tsExtremes package,  we start by loading the $\texttt{tsExtremes}$ package.

```{r setup}
library(tsExtremes)
```


The $\texttt{tsExtremes}$ package includes the rainfall dataset, containing daily rainfall records from nine different weather stations in France. The data and season of each record is available. To compute the tail-index and produce a Hill Plot of summer rainfall in Brest we use the function below.



```{r unbiased-hill, echo = T,fig.height = 4, fig.width = 9, fig.align = "center"}
h <- alphaestimator(rainfall$BREST[rainfall$SEASON=="SPRING"], plot=F , R0 = 100,  hill=T ,  k1 = 300 )
```


## References
